Analyzing DECREASE trials for extent of data fabrication

### 2017-03-02 submission/manuscript.tex
Revisions

\setlength{\droptitle}{-2em}
  \title{Analyzing DECREASE trials for extent of data fabrication}
  \pretitle{\vspace{\droptitle}\centering\huge}
  \posttitle{\par}
  \author{Chris HJ Hartgerink, Gerben ter Riet, Marleen Kemper}
  \preauthor{\centering\large\emph}
  \postauthor{\par}
  \predate{\centering\large\emph}
  \postdate{\par}
  \date{13 februari, 2017}

@@

The effect of beta-blockers on perioperative mortality in non-cardiac
surgery for patients with coronary artery disease (CAD) has been subject
of discussion\textsuperscript{1} due to findings of research
misconduct\textsuperscript{2--4}. Consequently, it appeared that the
related trials were central to claims of the effectiveness of
beta-blocker in decreasing perioperative
mortality\textsuperscript{5--7}. Perioperative mortality is the
deathrate of patients during the period of the surgical procedure, which
typically includes admission, anaesthesia, surgery, and recovery.
Meta-analyses that include these studies subject to research misconduct
conclude that beta-blockers decrease perioperative
mortality\textsuperscript{5--7} whereas a meta-analysis that excludes
the studies subject to research misconduct concludes that beta-blockers
increase perioperative mortality\textsuperscript{7}.

The trials subject to research misconduct were the Dutch DECREASE
trials\textsuperscript{2--4}. The committees that investigated the
integrity of the DECREASE studies reported that data fabrication was
likely but that the extent of the data fabrication remained
unclear\textsuperscript{2--4}. Given that the inclusion or exclusion of
these trials in meta-analyses is crucial in determining whether
beta-blockers are an effective treatment and considering that the latest
guidelines still recommend the usage of
beta-blockers\textsuperscript{8}, we aim to estimate the extent of data
fabrication in the DECREASE studies\textsuperscript{9,10} in order to
further the debate.

The reports on the integrity of the DECREASE trials primarily focused on
the provenance of the raw data and patient files (e.g., informed
consent, whether data corresponded to patient files) but did not
investigate the extent to which the DECREASE trials deviated from
comparable trials. The committee reports did not neglect statistical
evaluation; a statistical expert evaluated the applicability of forensic
statistical methods\textsuperscript{4} to evaluate results within one
trial. However, comparing across trials can be used as a forensic
statistical method as well. This method has previously been used to
monitor trial data quality or to test for potential data
anomalies\textsuperscript{11}. Moreover, this method has previously
proven to be effective in detecting data
fabrication\textsuperscript{12}. As such, evaluating the DECREASE trials
against comparable trials investigating the effectiveness of
beta-blockers with respect to perioperative mortality could prove
worthwhile.

To statistically investigate the evidence of data fabrication in the
results of the DECREASE studies\textsuperscript{9,10}, we took three
steps. First, we reproduced the findings from a 2014
meta-analysis\textsuperscript{7} that contained sufficient information
to estimate the deviation of the DECREASE-I and DECREASE-IV studies from
the remaining studies. Second, we evaluated the veracity of the
DECREASE-I and DECREASE-IV studies (i.e., the probability that these
study findings occurred assuming no data fabrication occurred). Third,
we reversed this assumption and assumed that data fabrication did occur
and estimate how many data points would have to be fabricated to
reproduce the results of the DECREASE-I and DECREASE-IV studies if the
remaining studies are regarded as estimating the true effect of
beta-blockers on perioperative mortality.

@@

\begin{verbatim}
## Warning: package 'latex2exp' was built under R version 3.2.5
\end{verbatim}

\begin{verbatim}
## Warning: package 'knitr' was built under R version 3.2.5
\end{verbatim}

We computed the log risk ratio (i.e., log RR) for each study and pooled
these using the \texttt{R} package \texttt{metafor}\textsuperscript{13}.
We estimated a weighted random-effects model using the restricted
maximum-likelihood estimator (i.e., \texttt{REML})\textsuperscript{14}
to estimate the variance of effects. We used the default weighting
procedure in the \texttt{metafor} package. When there was a zero-count
for a cell, .5 was added to each cell of that trial, as is common in
meta-analyses on risk- and odds ratios in order to deal with
computational artefacts\textsuperscript{15}. The 2014
meta-analysis\textsuperscript{7} did not specify the effect variance
estimate used; hence, minor discrepancies between our estimates and the
original estimates could be due to differences in the estimation
procedure.

@@

We analyzed the RRs to verify the original estimates (Figure 2 of the
2014 meta-analysis\textsuperscript{7}) and found that we were able to
reproduce the estimates for the different sets of studies. Bouri et al.
differentiated between the estimates from the non-DECREASE trials and
the DECREASE trials. We confirmed the effect size estimates and the
variance estimates for both the non-DECREASE- and the DECREASE trials,
save for some minor discrepancies due to the estimation method. Table 1
depicts the original and reproduced values for both sets of studies.

@@

Second, we meta-analyzed all studies combined, including a dummy
predictor for the DECREASE and non-DECREASE studies to reproduce results
presented in Figure 4 of the 2014 meta-analysis\textsuperscript{7}.
Surprisingly, our results showed more evidence for unequal subgroups
than the original meta-analysis\textsuperscript{7}; whereas the original
test for subgroup differences was \(\chi^2(1)=3.91,p=.05\), our analyses
indicated that \(\chi^2(1)=6.12,p=0.013\). Additionally, the original
analyses showed substantial heterogeneity (\(I^2=74.4\)\%) where we
found an \(I^2=0\)\%. Different heterogeneity estimators (e.g.,
\texttt{DerSimonian-Laird} instead of \texttt{REML}) did not resolve
this difference. We tried to clarify these discrepancies by e-mailing
the original authors, but did not receive a response.

@@

Based on the non-DECREASE estimates from Step 1, we estimated the
probability of obtaining the results in the DECREASE studies. To this
end, we assumed that the non-DECREASE studies provide a valid
representation of the true effect of beta-blockers on perioperative
mortality (similar to Bouri et al.\textsuperscript{7}; but see the
discussion). The estimated probability is also known as the veracity of
the data\textsuperscript{16}, which indicates the probability of the
observed data under a given true effect. We assumed that the
non-DECREASE studies estimate the true population distribution of
effects, not perturbed by publication bias due to statistical
(non)significance. Publication bias was assumed to not be a problem
because a substantial number of nonsignificant effects are included in
the dataset (9 of 11 nonsignificant).

@@

Based on the estimated mean log RR and its credible interval in the
non-DECREASE studies, we computed the probability of log RR in the
DECREASE trials. The estimates of the non-DECREASE studies were obtained
from Step 1, which include the estimated log RR (i.e., 0.25), and its
95\% credibility interval as provided by the package \texttt{metafor}
(i.e., {[}0.01; 0.484{]}. We assumed a normal distribution of population
effects with the estimated effect as the mean of the distribution. The
95\% credibility interval denotes the bounds of the normal distribution
that covers 95\% of the density, where the standard deviation is
calculated as the distance from the mean to either bound, divided by
1.96. This allows for an approximation of the population effect
distribution, as depicted in Figure 1.

@@

Based on the estimated effect distribution from the non-DECREASE trials,
we calculated the probability of each DECREASE trial result, or a more
extreme result. In other words, we computed the \emph{p}-value for the
hypothesis that the DECREASE studies occurred based on the information
available from the other trials.

@@

Figure 1 indicates that the DECREASE trials are highly unlikely under
the estimated effect distribution based on the non-DECREASE trials. More
specifically, DECREASE-I (or more extreme) has a probability of
\(2.9610213\times 10^{-53}\) (less than 1 in a quintillion) and
DECREASE-IV \(3.0974376\times 10^{-9}\) (3 in a billion), which
indicates they are unlikely to come from the same population effect
distribution as the non-DECREASE trials. Observing two of such extremely
unlikely results in this population effect distribution jointly is
highly improbable, \(9.1715786\times 10^{-62}\). This would therefore
indicate that the DECREASE trials are severely different from the
non-DECREASE studies.

Results from Step 1 indicated that no variance (i.e., homogeneity;
\(\tau^2=0\)) of the effects was observed; given the small number of
studies included (i.e., 9) this estimate is uncertain, however. We
conducted sensitivity analyses in order to see how dependent results are
on the specific heterogeneity estimate. The probability of observing
both DECREASE trials stays approximately below 1 out of 1000 until the
variance estimate is .25, and the probability stays approximately below
1 out of 100 until the variance estimate is .43 (see Figure 2). Further
research would be worthwhile to better estimate the variance in the
effects of beta-blockers on perioperative mortality and what moderators
affect the effect size (e.g., type of beta-blocker).

@@

We estimated the number of data points that would need to be fabricated
to arrive at the estimates from the DECREASE studies, given that the
non-DECREASE studies represent the true population effect distribution.
This assumes, in stark contrast to Step 2, that the DECREASE studies in
fact contain fabricated data. The estimates from Step 3 provide an
indication of the extent of potential data fabrication in the DECREASE
studies\textsuperscript{2--4,7}.

@@

In order to estimate the number of fabricated data points, we estimated
the population effect distribution of mortality (in log odds) per
condition in both the DECREASE- and non-DECREASE studies. For each of
these, we ran a meta-analysis applying the same, verified methods used
in Step 1. These separate analyses were necessary, considering that Step
1 only provided estimates of the overall risk ratio but not the odds
conditional on exposure (beta-blocker or control) and trial set
(DECREASE- versus non-DECREASE). This resulted in four meta-analytic
mortality estimates with an accompanying effect variance. Throughout the
simulations, we use the point estimates to simulate genuine- and
fabricated data but supplement this by using the more uncertain
distribution estimates as sensitivity analyses.

We applied the inversion method to estimate the number of fabricated
data points in the DECREASE-I and DECREASE-IV
trials\textsuperscript{17}. With the inversion method, we iteratively
hypothesized that \emph{X} out of \emph{N} data points were fabricated
(i.e., \(X={0, 1, ..., N}\)). For each combination of \emph{X} and
trial, we simulated 10,000 datasets. Each simulated dataset contained
\emph{X} fabricated data points and \emph{N-X} genuine data points. For
each simulated dataset (exact simulation procedure in next paragraph),
we determined the likelihood of the results with

@@

where \(\pi_{E}\) indicates the mortality rate of the experimental
condition as drawn from the meta-analytic effect distribution
(\(\pi_{C}\) indicates the same for the control condition). The
likelihood was computed under both the fabricated effect estimates
(i.e., \(L_{fabricated}\)) and the genuine data (i.e., \(L_{genuine}\)).
Table 2 indicates which cell sizes the various \(n_{XX}\) refer to
within the (simulated) data. Subsequently, we compared the likelihoods
to determine whether the simulated data were more likely to arise from
the genuine trials (\(L_{genuine}\)) or from the fabricated trials
(\(L_{fabricated}\)). Note that using the likelihoods for this is a
minor deviation from the preregistration, where we initially planned on
using \(p\)-value comparisons
(\href{https://osf.io/vnmzc}{osf.io/vnmzc}).

@@

We computed \(p_F\) for all \emph{X} out of \emph{N} fabricated
datapoints in 10,000 randomly generated datsets, which were generated in
three steps. For each dataset we:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  Sampled (without replacement and across conditions) \emph{X}
  fictitious participants that would be the result of data fabrication.
\item
  Determined the population mortality rate for each condition (i.e., for
  each cell as in Table 3). The meta-analytic point estimate was used or
  a population effect was randomly drawn from the meta-analytic effect
  distribution.
\item
  Simulated the number of deaths for the different conditions using a
  binomial distribution based on the mortality rate as determined in 2,
  resulting in the cell counts as in Table 3.
\end{enumerate}

Based on the meta-analytic effect from 2 and the cell sizes from 3, we
computed the likelihoods of \(L_{fabricated}\) and \(L_{genuine}\) using
Equation \ref{eq1}. As mentioned before, we computed \(p_F\) as the
probability that the data were more likely under the estimates resulting
from the (allegedly) fabricated data (i.e., the DECREASE trials) than
under the estimates resulting from the genuine data (i.e., the
non-DECREASE trials; \(p_F=P(L_{fabricated}>L_{genuine})\)).

@@

For DECREASE-I, the 95\% confidence interval for the estimated number of
fabricated data points is {[}0 - 112{]} or {[}3 - 112{]} when based on a
point estimate or a more uncertain distribution estimate, respectively.
The left column of Figure 3 depicts the \(p_F\) per \emph{X} fabricated
data points (top panel) and the bounds of the confidence interval when
the degree of confidence is altered (lower panel). Staying clearly
between the dotted lines in the top panel, depicting the 95\% CI (top:
.975; bottom: .025), it becomes apparent that the degree of uncertainty
is too high to make any estimates about fabricated data points with a
high degree of confidence. This is partly due to the small sample size
(i.e., \(N=112\)) and the availability of just the summary results. Only
when the degree of confidence is lowered to around 75\% does the
interval not span the entire sample size. As such, based on the summary
results, little can be said about the extent of the data fabrication
that occurred in the DECREASE-I trial, affirming the conclusions of the
original committee report\textsuperscript{4}.

For DECREASE-IV, the 95\% confidence interval for the estimated number
of fabricated data points is {[}8 - 1059{]} or {[}7 - 1066{]} when based
on a point estimate or a more uncertain distribution estimate,
respectively. The minor difference between the estimates indicates that
there is a high degree of confidence that data fabrication did occur
based on the difference of the results alone. Nonetheless, the range of
potentially fabricated data points is still estimated at approximately
1000; this indicates that the summary results are insufficient to
provide more than an estimated lowerbound. This does indicate that it is
possible not all data were fabricated (i.e., \(N=1066\)), increasing the
importance of well-documented data provenance to discern between genuine
and falsified data. This affirms the conclusion of the scientific
integrity committee that the results of the DECREASE-IV trial are
``scientifically incorrect'' (p.11)\textsuperscript{3}.

The results also highlight that, despite the lack of availability of raw
data, summary results from larger samples result in more certainty about
the estimated number of fabricated data points using the inversion
method. This is due to decreased standard errors at the population level
of the effects, providing less variation in the simulations and
resulting in higher sensitivity. However, this also highlights that in
order to prevent detection it would be in the fabricators' interest to
fabricate small studies (assuming the fabricator wants to remain
undetected). This adds another factor to the unreliability of studies
with small sample sizes.

@@

We note that the variance in the non-DECREASE studies was estimated at
0, indicating that there would supposedly be a fixed-effect of
beta-blockers on perioperative mortality despite differences in type of
beta-blockers administered and dosage/duration of the beta-blockers.
This homogeneity is not excessive, considering that the
\emph{Q}-statistic is sufficiently large to begin with (i.e.,
5).\textsuperscript{18} Nonetheless, the original non-DECREASE trials
show substantive variation in the treatment, for example the usage of
different beta-blockers (i.e., atenolol, propanolol, bisoprolol,
metoprolol) in different (standardized) dosages, which have been argued
to affect effectiveness of the interventions\textsuperscript{19}. As
such, the lack of heterogeneity can be genuine and indicative of no
difference between these different beta-blockers and their
dosage/duration. Nonetheless, it could also be due to uncertainty in the
estimate of heterogeneity \(\tau^2\) due to the small set of studies
included.

However, due to various types of beta-blockers, evidence might be
moderated\textsuperscript{19}.

@@

The ESC/ESA guidelines from 2014{[}@{]} were published in October 2014,
whereas the meta-analysis indicating a reversal of the effectiveness of
beta-blockers was published online at the end of July 2013{[}@{]}. The
committee revising the ESC/ESA guidelines stated: ``The respective
writing committees independently performed their literature review and
analysis, and then developed their
recommendations.''\textsuperscript{20} Regardless, the revisions made
``recommend continuation of beta-blocker therapy in the perioperative
period in patients currently receiving this medication'' and do not
recommend the ``initiation of beta-blockers in patients undergoing
low-risk surgery''\textsuperscript{21}. Review pieces note no
substantial clinical change with respect to
beta-blockers.\textsuperscript{22,23}

The revisions to the ESC/ESA guidelines do not discourage use of
beta-blockers, despite the obvious lack of evidence for systematic
evidence of the effectiveness of beta-blockers to reduce perioperative
mortality in trials. If anything, the trials show that insufficient
evidence has been collected. This becomes apparent when the estimated
effect for all non-DECREASE trials is inspected; its \emph{p}-value is
.04, which is insufficient evidence and is actually more likely when
there is no effect in of beta-blockers in reality\textsuperscript{24}

@@

\hypertarget{ref-dunkelgrun2009}{}
9. Dunkelgrun M, Boersma E, Schouten O, et al. Bisoprolol and
fluvastatin for the reduction of perioperative cardiac mortality and
myocardial infarction in intermediate-risk patients undergoing
noncardiovascular surgery: A randomized controlled trial (DECREASE-IV).
\emph{Annals of surgery}. 2009;249(6):921-926.
doi:\href{https://doi.org/10.1097/SLA.0b013e3181a77d00}{10.1097/SLA.0b013e3181a77d00}.

>>>>

\setlength{\droptitle}{-2em}
  \title{Analyzing DECREASE trials for extent of data fabrication}
  \pretitle{\vspace{\droptitle}\centering\huge}
  \posttitle{\par}
  \author{Chris HJ Hartgerink, Gerben ter Riet, Marleen Kemper}
  \preauthor{\centering\large\emph}
  \postauthor{\par}
  \predate{\centering\large\emph}
  \postdate{\par}
  \date{02 March, 2017}

@@

The effect of beta-blockers on perioperative mortality in non-cardiac
surgery for patients with coronary artery disease (CAD) has been
controversial\textsuperscript{1} due to findings of research misconduct
in two related clinical trials\textsuperscript{2--4}. Three
meta-analyses that included the trials subject to research misconduct
concluded that beta-blockers decrease perioperative
mortality\textsuperscript{5--7} whereas a meta-analysis that excluded
the suspect trials concluded that beta-blockers increase perioperative
mortality\textsuperscript{7}. In these studies, perioperative mortality
was defined as the deathrate of patients during the period of the
surgical procedure, which typically includes admission, anaesthesia,
surgery, and recovery.

The trials subject to research misconduct were the Dutch DECREASE-I and
DECREASE-IV trials\textsuperscript{2--4}. The committees that
investigated the integrity of the DECREASE trials reported that data
fabrication was likely but that the extent of the data fabrication
remained unclear\textsuperscript{2--4}. Moreover, the latest guidelines
still recommend the usage of beta-blockers in the perioperative period
in certain cases\textsuperscript{8,9}. Considering these consequences,
we aim to estimate the extent of data fabrication in the DECREASE
studies\textsuperscript{10,11} to further the debate on using
beta-blockers in the perioperative period of CAD patients undergoing
non-cardiac surgery.

The reports on the integrity of the DECREASE trials primarily focused on
the provenance of the raw data but did not investigate the extent to
which the DECREASE trials deviated from comparable trials. The
provenance is primarily concerned with the origins of the data,
verifying things such as (but not limited to) the informed consent and
whether data corresponded to patient files. The committee reports did
not neglect statistical evaluation however: a statistical expert
evaluated the applicability of forensic statistical
methods\textsuperscript{4} to evaluate results of trials separately
(i.e., DECREASE-I, DECREASE-IV). Nonetheless, comparing across trials is
a method that has previously been used to monitor trial data quality or
to test for potential data anomalies\textsuperscript{12}. Moreover, this
method has previously proven to be effective in detecting data
fabrication\textsuperscript{13}. Evaluating the DECREASE trials to other
published trials studying the effectiveness of beta-blockers with
respect to perioperative mortality could prove informative of the
potential extent of the fabrication in the DECREASE trials.

However, the effectiveness of perioperative beta-blockade is obfuscated
by both trials afflicted by data fabrication and the type of
beta-blocker used in the trial. Various kinds of beta-blockers could
vary in their pharmalogical effectiveness with respect to perioperative
mortality, and anomalies in the effectiveness of the DECREASE trials
might be due to differences in the beta-blockers used {[}@{]} and not
purely due to data fabrication (given not all data points can be
considered fabricated at this point).

To statistically investigate the evidence of data fabrication in the
DECREASE studies\textsuperscript{10,11}, we took three steps. First, we
reproduced the findings from a 2014 meta-analysis\textsuperscript{7}
that contained sufficient information to estimate the deviation of the
DECREASE trials from other published trials on beta-blockers. We also
include type of beta-blocker to inspect whether this is predictive of
the effect of beta-blockers on perioperative mortality. Second, we
evaluated the the probability that the DECREASE trial results occurred
assuming no data fabrication. Third, we reversed this assumption and
assumed that data fabrication did occur and estimated how many data
points would have to be fabricated to reproduce the results of the
DECREASE trials, if the other published trials are regarded as
estimating the true effect of beta-blockers on perioperative mortality
in CAD patients undergoing non-cardiac surgery.

@@

We computed the log risk ratio (i.e., log RR) for each study and pooled
these using the \texttt{R} package \texttt{metafor}\textsuperscript{14}.
We estimated a weighted random-effects model using the restricted
maximum-likelihood estimator (i.e., \texttt{REML})\textsuperscript{15}
to estimate the variance of effects. We used the default weighting
procedure in the \texttt{metafor} package. When there was a zero-count
for a cell (e.g., zero events), 0.5 was added to that cell, as is common
in meta-analyses on risk- and odds ratios in order to prevent
computational artefacts\textsuperscript{16}. The 2014
meta-analysis\textsuperscript{7} did not specify the variance estimate
used; hence, minor discrepancies between our estimates and the original
estimates could be due to differences in the estimation procedure.

@@

We were able to closely reproduce the estimates for the different sets
of studies (Figure 2 of the 2014 meta-analysis\textsuperscript{7}).
Bouri et al. differentiated between the estimates from the non-DECREASE
trials and the DECREASE trials. We confirmed the effect size estimates
and the variance estimates for both the non-DECREASE- and the DECREASE
trials, save for some minor discrepancies due to the estimation method.
Table 1 depicts the original and reproduced values for both sets of
studies.

@@

Second, we meta-analyzed all studies combined, including a dummy
predictor for the DECREASE and non-DECREASE studies to reproduce results
presented in Figure 4 of the 2014 meta-analysis\textsuperscript{7}.
Surprisingly, our results showed stronger evidence against equal
subgroups than the original meta-analysis\textsuperscript{7} (original:
\(\chi^2(1)=3.91,p=.05\), reproduced \(\chi^2(1)=6.12,p=0.013\)).
Additionally, the original analyses showed substantial residual
heterogeneity (\(I^2=74.4\)\%) whereas we found no residual heterogeneit
(\$I\^{}2=\$0\%). Different variance estimates (e.g.,
\texttt{DerSimonian-Laird} instead of \texttt{REML}) did not resolve
this difference. We tried to clarify these discrepancies by e-mailing
the original authors, but did not receive a response. Nonetheless, the
broad strokes of the meta-regression confirmed that the DECREASE trials
were the determining predictor for the effectiveness of beta-blockers
(including DECREASE: \(RR=0.496\); excluding DECREASE: \(RR=1.28\)).

Additionally, and exploratively, we evaluated the predictive effect of
the type of beta-blocker used in the trials. The DECREASE trials remain
predictive of decreased mortality (\(RR=0.496\)), whereas the
non-DECREASE trials provide tentative, but uncertain, evidence that
atenolol results in lower mortality (\(RR=0.751\)). Nonetheless, for
other beta-blockers in the non-DECREASE trials, there is still tentative
and uncertain evidence that beta-blockers increase mortality
(bisoprolol: \(RR=2.973\); metoprolol: \(RR=1.307\); propanolol:
\(RR=2.041\)). Table 3 shows the meta-regression results in full.

\begin{longtable}[]{@{}lll@{}}
\caption{Meta-regression results for log(RR), including dummy predictors
for DECREASE trials (reference: non-DECREASE trials) and type of
beta-blockers used in the trial (reference: atenolol).}\tabularnewline
\toprule
& Estimate & 95\% CI\tabularnewline
Intercept & -0.287 & -1.317; 0.743\tabularnewline
Non-DECREASE & &\tabularnewline
DECREASE & -1.792 & -5.082; 1.498\tabularnewline
Atenolol & &\tabularnewline
Bisoprolol & 1.376 & -1.996; 4.749\tabularnewline
Metoprolol & 0.554 & -0.504; 1.613\tabularnewline
Propanolol & 1 & -1.642; 3.643\tabularnewline
\bottomrule
\end{longtable}

@@

Based on the non-DECREASE estimates from Step 1, we estimated the
probability of obtaining the results in the DECREASE trials. To this
end, we assumed that the non-DECREASE trials provide a valid
representation of the true effect of beta-blockers on perioperative
mortality (similar to Bouri et al.\textsuperscript{7}; but see the
discussion). The estimated probability is also known as the veracity of
the data\textsuperscript{17}, which indicates the probability of the
observed data under a given true effect. We assumed that the
non-DECREASE studies estimated the true effect distribution of
perioperative beta-blockade on mortality, not perturbed by publication
bias due to statistical (non)significance. Publication bias was assumed
to not be a problem because a substantial number of nonsignificant
effects are included in the dataset (9 of 11 nonsignificant results).

@@

Based on the estimated mean log RR and its credible interval in the
non-DECREASE studies, we computed the probability of the observed log RR
in the DECREASE trials. The estimates of the non-DECREASE studies were
obtained from Step 1, which include the estimated log RR (i.e., 0.25),
and its 95\% credibility interval as provided by the package
\texttt{metafor} (i.e., {[}0.01; 0.484{]}. The meta-analysis model
assumes a normal distribution of population effects with the estimated
effect as the mean of the distribution. The 95\% credibility interval
denotes the bounds of the normal distribution that covers 95\% of the
density, where the standard deviation is calculated as the distance from
the mean to either bound, divided by 1.96. This allows for an
approximation of the population effect distribution, as depicted in
Figure 1.

@@

Based on the estimated effect distribution from the non-DECREASE trials,
we calculated the probability of each DECREASE trial result, or a more
extreme result. In other words, we computed the \emph{p}-value for the
hypothesis that the DECREASE studies occurred, assuming the information
available from the other trials is informative of the true population
effect.

@@

Figure 1 indicates that the DECREASE trials are highly unlikely under
the estimated effect distribution based on the non-DECREASE trials. More
specifically, the results from DECREASE-I (or more extreme) have a
probability of \(0\) (less than 1 in a quintillion) and the results from
DECREASE-IV have a probability of \(0\) (3 in a billion). This indicates
the DECREASE trial results are unlikely to have come from the same
population effect distribution as the non-DECREASE trials. Moreover,
observing two of such extremely unlikely results jointly, as in the
DECREASE trials, is highly improbable, \(0\). This would therefore
indicate that the DECREASE trials are severely different from the
non-DECREASE trials.

Results from Step 1 indicated that no between-study variance (i.e.,
homogeneity; \(\tau^2=0\)) of the effects was observed; given the small
number of studies included (i.e., 9) this estimate is uncertain,
however. We conducted sensitivity analyses to see how dependent results
are on the heterogeneity estimate. The probability of observing the
DECREASE trials stays approximately below 1 out of 1000 until the
variance estimate is 0.25; the probability stays approximately below 1
out of 100 until the variance estimate is 0.43 (see Figure 2). To put
these numbers into context, a variance of 0.25 would suggest that
results of perioperative beta-blockade vary substantially even if
perioperative beta-blockade has no effect whatsoever (RRs between 0.779
and 1.284 in 64\% of the cases) all due to contextual circumstances of
the study.

@@

We estimated the number of data points that would need to be fabricated
to arrive at the estimates from the DECREASE trials, given that the
non-DECREASE trials represent the true effect of perioperative
beta-blockade. In contrast to Step 2 this assumes that the DECREASE
trials might in fact contain fabricated data. The estimates from Step 3
provide an indication of the extent of potential data fabrication in the
DECREASE studies\textsuperscript{2--4,7}.

@@

In order to estimate the number of fabricated data points, we first
estimated the effect of perioperative beta-blockade on mortality (in log
odds) in each trial arm. In total there are four trial arms: one per
condition (beta-blocker or control) per trial type (DECREASE- and
non-DECREASE trials). For each of these trial arms, we ran a
meta-analysis applying the same methods used in Step 1. For each of the
two DECREASE trials separately and the non-DECREASE trials combined,
this resulted in four meta-analytic mortality estimates with
corresponding effect variances (see Table 3). Throughout the
simulations, we used the point estimates (i.e., fixed effect) to
simulate genuine- and fabricated data, but supplemented this by using
the more uncertain distribution estimates (i.e., random effects) as
sensitivity analyses.

We applied the inversion method to estimate the number of fabricated
data points in the DECREASE trials\textsuperscript{18}. The inversion
method iteratively hypothesizes that \emph{X} out of \emph{N} data
points were fabricated (i.e., \(X={0, 1, ..., N}\)). For each
combination of \emph{X} and trial, we simulated 10000 datasets. Each
simulated dataset contained \emph{X} fabricated data points and
\emph{N-X} genuine data points. For each simulated dataset (exact
simulation procedure in the next paragraph), we determined the
likelihood of the results with

@@

where \(\pi_{E}\) indicates the mortality rate in the beta-blocker
condition as drawn from the meta-analytic effect distribution
(\(\pi_{C}\) indicates the mortality rate in the control condition). The
likelihood was computed under both the fabricated effect estimates
(i.e., \(L_{fabricated}\)) and the genuine data (i.e., \(L_{genuine}\)).
Table 3 indicates which cell sizes the various \(n_{XX}\) refer to
within the (simulated) data. After computing the likelihoods, we
compared them to determine whether the simulated data were more likely
to arise from the genuine trials (\(L_{genuine}>L_{fabricated}\)) or
from the fabricated trials (\(L_{fabricated}>L_{genuine}\)). Note that
comparing the likelihoods is a minor deviation from the preregistration,
where we initially planned on using \(p\)-value comparisons
(\href{https://osf.io/vnmzc}{osf.io/vnmzc}).

@@

We computed \(p_F\) for all \emph{X} out of \emph{N} fabricated
datapoints in 10000 randomly generated datasets, which were generated in
three steps. For each dataset we:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  Sampled (across conditions, without replacement) \emph{X} fictitious
  participants that would be the result of data fabrication.
\item
  Determined the population mortality rate for each condition (i.e., for
  each cell as in Table 3). The meta-analytic point estimate was used or
  a population effect was randomly drawn from the meta-analytic effect
  distribution.
\item
  Simulated the number of deaths for the different conditions using a
  binomial distribution based on the mortality rate as determined in 2,
  resulting in the cell counts as in Table 3.
\end{enumerate}

Based on the meta-analytic effect from 2 and the cell sizes from 3, we
computed the likelihoods \(L_{fabricated}\) and \(L_{genuine}\) using
Equation \ref{eq1}. As mentioned before, we computed \(p_F\), which
indicated the probability that the data are more likely under the
estimates resulting from the (allegedly) fabricated data (i.e., the
DECREASE trials) than under the estimates resulting from the genuine
data (i.e., the non-DECREASE trials;
\(p_F=P(L_{fabricated}>L_{genuine})\)).

@@

For DECREASE-I (\$N=112), the 95\% confidence interval for the estimated
number of fabricated data points is {[}0 - 112{]} or {[}2 - 112{]} when
based on a point estimate or a more uncertain distribution estimate,
respectively. The left column of Figure 3 depicts the \(p_F\) per
\emph{X} fabricated data points (top panel) and the bounds of the
confidence interval when the degree of confidence is altered (lower
panel). Staying clearly between the dotted lines in the top panel,
depicting the 95\% CI (top: .975; bottom: .025), it becomes apparent
that the degree of uncertainty is too high to make any reasonable
estimates about the number of fabricated data points with sufficient
confidence. This is partly due to the small sample size of the
DECREASE-I trial (i.e., \(N=112\)) and the availability of just the
summary results. Only when the degree of confidence is lowered to around
75\% does the interval not span the entire sample size. As such, based
on the summary results, little can be said about the extent of the data
fabrication that occurred in the DECREASE-I trial, affirming the
conclusions of the original committee report\textsuperscript{4}.

For DECREASE-IV (\$N=1066), the 95\% confidence interval for the
estimated number of fabricated data points is {[}0 - 1053{]} or {[}8 -
1066{]} when based on a point estimate or a more uncertain distribution
estimate, respectively. The relatively minor difference between the
estimates indicates that there is a high degree of confidence that data
fabrication did occur based on the difference of the trial results
alone. Nonetheless, the range of potentially fabricated data points is
still estimated at approximately 1000; this indicates that the summary
results are insufficient to provide more than an estimated lowerbound.
This indicates that it is possible not all data were fabricated (i.e.,
\(N=1066\)), increasing the importance of well-documented data
provenance to discern between genuine and falsified data. This affirms
the conclusion of the scientific integrity committee that the results of
the DECREASE-IV trial are ``scientifically incorrect''
(p.11)\textsuperscript{3}.

The results also highlight that, despite the lack of availability of the
raw data, summary results from larger samples allow estimation of the
number of fabricated data points when similar trials are available.
Moreover, larger trials result in relatively more certainty (e.g.,
DECREASE-IV) about the estimated number of fabricated data points using
the inversion method compared to smaller trials (e.g., DECREASE-I),
although much residual uncertainty remains. This increased certainty is
due to decreased standard errors at the population level of the
estimated effects, resulting in higher sensitivity to data anomalies.
Nonetheless, substantially less information is available in just the
summary results, and raw data availability would improve results
verification and the detection of potential anomalies. These results
also highlight that in order to prevent detection, it would be in the
fabricators' interest to fabricate small studies even if raw data are
hidden away (assuming the fabricator wants to remain undetected).

@@

We note that the variance in the non-DECREASE studies was estimated at
0, indicating that there would supposedly be a fixed-effect of
beta-blockers on perioperative mortality despite differences in type of
beta-blockers administered and dosage/duration of the beta-blockers.
This homogeneity is not excessive, considering that the
\emph{Q}-statistic is sufficiently large to begin with (i.e.,
5).\textsuperscript{19} Nonetheless, the original non-DECREASE trials
show substantive variation in the treatment, for example the usage of
different beta-blockers (i.e., atenolol, propanolol, bisoprolol,
metoprolol) in different (standardized) dosages, which have been argued
to affect effectiveness of the interventions\textsuperscript{20}. As
such, the lack of heterogeneity can be genuine and indicative of no
difference between these different beta-blockers and their
dosage/duration. Nonetheless, it could also be due to uncertainty in the
estimate of heterogeneity \(\tau^2\) due to the small set of studies
included.

However, due to various types of beta-blockers, evidence might be
moderated\textsuperscript{20}.

@@

The ESC/ESA guidelines from 2014{[}@{]} were published in October 2014,
whereas the meta-analysis indicating a reversal of the effectiveness of
beta-blockers was published online at the end of July 2013{[}@{]}. The
committee revising the ESC/ESA guidelines stated: ``The respective
writing committees independently performed their literature review and
analysis, and then developed their
recommendations.''\textsuperscript{21} Regardless, the revisions made
``recommend continuation of beta-blocker therapy in the perioperative
period in patients currently receiving this medication'' and do not
recommend the ``initiation of beta-blockers in patients undergoing
low-risk surgery''\textsuperscript{22}. Review pieces note no
substantial clinical change with respect to
beta-blockers.\textsuperscript{23,24}

The revisions to the ESC/ESA guidelines do not discourage use of
beta-blockers, despite the obvious lack of evidence for systematic
evidence of the effectiveness of beta-blockers to reduce perioperative
mortality in trials. If anything, the trials show that insufficient
evidence has been collected. This becomes apparent when the estimated
effect for all non-DECREASE trials is inspected; its \emph{p}-value is
.04, which is insufficient evidence and is actually more likely when
there is no effect in of beta-blockers in reality\textsuperscript{25}

@@

\hypertarget{ref-Fleishere278}{}
9. Fleisher LA, Fleischmann KE, Auerbach AD, et al. 2014 acc/aha
guideline on perioperative cardiovascular evaluation and management of
patients undergoing noncardiac surgery. \emph{Circulation}.
2014;130(24):e278-e333.
doi:\href{https://doi.org/10.1161/CIR.0000000000000106}{10.1161/CIR.0000000000000106}.
